---
title: "TableMt_2012"
author: "D. Ackerly, J. Slingsby"
date: "February 14, 2016"
output: html_document
---

In 2012, we placed 96 dataloggers recording temperature and humidity across Table Mt. and Silvermine. The loggers were stratified across gradients of elevation, slope, exposure and hillslope position (hilltop, valleybottom), and distance from both False Bay and the Atlantic. In previous analyses, the data has been analyzed to extract diurnal Tmin, Tmax, VPmax (maximum daily vapor pressure), and RHsat.hrs (hours per day with relative humidity above 95%).

All of these analyses and the summary data are available at: 
https://github.com/dackerly/table-mt-microclimate-2012.git

General summary statistics about the sites and daily weather values can be found in 'manuscript-summary-stats.r'. 

The goal of these analyses is to examine spatial variation in daily weather values, running models for each day through 2012. The initial question is whether the important spatial variables vary with season, but as shown below a key result is enormous day to day variation in the ability to explain spatial variation in weather variables. I'm still searching for explanations in regional weather or other factors to understand why the models differ so much from day to day.

```{r}
#----------------#
# LOAD LIBRARIES #
#----------------#

rm(list=ls())
library(raster)
library(RColorBrewer)

#----------------#
# LOAD DATA      #
#----------------#
```
'meta' contains metadata on the 96 sites. 'topo10' and 'topo30' have a series of topographic features sampled from 10m and 30m dem rasters, respectively, based on lat/long positions of each site. SiteID is the unique name for each site. 

Some of the variables in 'meta':
elevation, slope, and aspect: measured in the field - these are generally not used, and instead we use a consistent set of values taken from the 10m dem.
use4Analyses, use4ClimateSummaries: sites excluded due to incomplete data (see ms for criteria)

Variables in topo10 and topo30:
d2at, d2fb, d2cs: distance to atlantic, distance to false bay, distance to coast = min(d2at,d2fb)
rad080, rad172, rad355: solar radiation at summer and winter solstice, and equinox. Not generally used as we have diurnal radiation at each site
thl0, thl315, thl337: Topographic heat load as defined by McCune and Keon 2002 JVS, with the axis for calculation of the 'folded aspect' (and thus the maximum heat load) at 0, 337 or 315 degrees. The logic is that in the S Hem, NW facing slopes would be hotter than N facing, due to higher afternoon temperatures.
plow050, 125, 250, 500: Percent lower pixels - the percent of pixels surrounding a focal pixel with a lower elevation, in radii of 50, 125, 250 or 500 m. A measure of potential for cold air to flow away from a spot. Higher values indicate hilltops
tpi050, 125, 250, 500: Elevation of a focal pixel - mean elevation in the specified radius. Another measure of hillslope position, with positive values for hilltops, negative for valley bottoms.
```{r}
meta <- read.csv('data/csv_masters/location_meta.csv',as.is=T)
dim(meta)
head(meta)

topo10 <- read.csv('data/csv_masters/topo10.csv',as.is=T)
dim(topo10)
names(topo10)
 
topo30 <- read.csv('data/csv_masters/topo30.csv',as.is=T)
dim(topo30)
names(topo30)
```

'dw' has all daily weather values for all sites and days. The dlySummary contains summaries of weather data on each day - average weather across sites. dlySol has diurnal radiation at each site for each day of the year, from Adam Wilson's analyses on a 30m dem.
```{r}

dw <- read.csv('data/csv_masters/2012daily.csv',as.is=T)
head(dw)

dlySummary <- read.csv('data/csv_outfiles/dlySummary.csv',as.is=T)
head(dlySummary)

dlySol <- readRDS('data/Rdata/dlySol.Rdata')
dlySol[1:6,1:6]

```

Before fitting any spatial models of weather variation, we want to check for correlations and spatial autocorrelation of predictors. Based on the topo10 variables, here are all variables with pairwise correlations R2 > 0.5. Commented code can be run to see more detail. As expected, radiation load variables are mostly correlated, and hillslope position variables are correlated. Below, we run alternative models swapping these in and out, but not using them in the same model.

```{r}

# cor(topo10[,-c(1:2)],use='pair')
# pairs(topo10[,-c(1:2)])
# names(topo10)
# 
## find correlations > 0.5
i <- 3; j <- 4;
for (i in 3:(ncol(topo10)-1)) for (j in (i+1):ncol(topo10)) {
    rr <- cor(topo10[,i],topo10[,j],use='pair')
    if (abs(rr)>sqrt(0.5)) print(c(names(topo10)[c(i,j)],round(rr,3)))
}
# 
# # distance from Atlantic and False Bay, borderline
# # Ran some models swapping these two variables in and out, and found false bay on average increased R2 more, and overall model R2 are significantly correlated so when one works, either works overall. Based on this, all models use Distance to False Bay.
# cor(topo10[,c('d2at','d2fb')],use='pair') # R = -0.65
# 
# # equinox and winter solstice radiation
# cor(topo10[,c('rad080','rad172')],use='pair') # R = 0.95
# 
# # all tpi variables with each other across scales, and plow with each other across scales; elevation weakly correlated with tpi and plow
# cor(topo10[,c('elevation','tpi050','tpi125','tpi250','tpi500','plow050','plow125','plow250','plow500')],use='pair') # R >= 0.89
# cor(topo30[,c('elevation','tpi050','tpi125','tpi250','tpi500','plow050','plow125','plow250','plow500')],use='pair') # R >= 0.89
# 
```

How does the Topographic Heat Load index, calculated with different aspect orientations, compare to daily radiation load from a solar model? Maximum values are near the equinoxes, with a minimum at summer solstice and a dip at winter solstice. However, I don't understand with the thl315 correlations are lower than the others, and wonder if the calculation was done right. Perhaps don't use that for now.
```{r}
# # How does daily solar radiation compare to topographic heat load
# # correlation max at 0.7 on day 48 (Feb 17) and around day 295 (Oct 21)
# # minimum mid-summer solstice
# head(dlySol)
# head(topo10)
solVthl <- c()
i=1
for (i in 1:366) solVthl[i] <- cor(topo10$thl0,dlySol[,i],use='pair')
plot(1:366,solVthl,type='l')   

for (i in 1:366) solVthl[i] <- cor(topo10$thl315,dlySol[,i],use='pair')
lines(1:366,solVthl,col='red')

for (i in 1:366) solVthl[i] <- cor(topo10$thl337,dlySol[,i],use='pair')
lines(1:366,solVthl,col='blue')   

```

The following shows scatterplots of pairwise differences matrices for spatial distance vs. each environmental factor, and the accompanying Mantel test. The plots that are commented out are not significant for spatial correlation, and plots that are significant are illustrated. The random string in the box is an identifier to link this code to the results in the manuscript. Site 32 (Kirstenbosch Race Track) is removed from analysis.

As expected, distance from ocean variables are strongly spatially correlated. This raises questions about their interpretation in the models, but they are generally their for the purpose of capturing spatial gradients, so that could be okay. Elevation and slope are also spatially structured, though the scatterplots suggests less concern. Exposure and hillslope positions are not correlated, indicating we did a good job stratifying across the region.

```{r}
# 
# #--------------------------------------------#
# # SPATIAL VARIATION IN TOPOGRAPHIC VARIABLES #
# # a6STycTJVjqM8vRd8V8h                       #
# #--------------------------------------------#
library(ade4)
# check matrices are aligned
all(meta$siteID==topo10$siteID)

#head(meta)
sdist <- dist(meta[-32,c('UTM.east','UTM.north')])
hist(sdist,main='Histogram of distances between sites (m)')

selev <- dist(meta$elevation[-32])
plot(sdist,selev)
mantel.rtest(sdist,selev,nrepet=999)

sd2at <- dist(topo10$d2at[-32])
plot(sdist,sd2at)
mantel.rtest(sdist,sd2at,nrepet=999)

sd2fb <- dist(topo10$d2fb[-32])
plot(sdist,sd2fb)
mantel.rtest(sdist,sd2fb,nrepet=999)

plot(selev,sd2fb)
mantel.rtest(selev,sd2fb,nrepet=999)

sd2cs <- dist(topo10$d2cs[-32])
plot(sdist,sd2cs)
mantel.rtest(sdist,sd2cs,nrepet=999)

dslope <- dist(topo10$slope[-32])
plot(sdist,dslope)
mantel.rtest(sdist,dslope,nrepet=999)

dtpi50 <- dist(topo10$tpi050[-32])
#plot(sdist,dtpi50)
mantel.rtest(sdist,dtpi50,nrepet=999)

drad080 <- dist(topo10$rad080[-32])
#plot(sdist,drad080)
mantel.rtest(sdist,drad080,nrepet=999)

dthl <- dist(topo10$thl0[-32])
#plot(sdist,dthl)
mantel.rtest(sdist,dthl,nrepet=999)
```

```{r}
# 
# #-----------------------#
# # LOAD DLY WEATHER DATA #
# # Leaving commented out for now, as this has been replaced by 'dw' data.frame #
# #-----------------------#
# 
# cbday <- readRDS('data/Rdata/cbday.Rdata')
# length(cbday)
# head(cbday[[1]])
# dim(cbday[[1]])
# 
# #-------------------------------------------------#
# # MAKE 3 CHAR DAYS VAR, for dlySol variable names #
# #-------------------------------------------------#
days <- as.character(1:366)
days[nchar(days)==1] <- paste('00',days[nchar(days)==1],sep='')
days[nchar(days)==2] <- paste('0',days[nchar(days)==2],sep='')
```

Okay, the modeling strategy is a bit brute force now!

```{R}
# #-------------------------------#
# # DAILY WEATHER MODELS          #
# # OHkNt68RXvbTk4jHfczb          #
# #-------------------------------#
# 
# # Run through daily models testing all combinations of elevation, radiation, hillslope, and regional position terms
# radterm <- c('dsol','thl315','rad080','rad172','rad355')
# hillterm <- c('tpi050','tpi125','tpi250','tpi500','plow050','plow125','plow250','plow500')
# regterm <- c('d2fb','d2at','d2cs')
# slterm <- 'slope'
# nModels <- length(radterm)*length(hillterm)*length(regterm)
# 
# modterms <- matrix(NA,120,5)
# modnums <- matrix(NA,120,5)
# n <- 0
# for (i in 1:length(radterm)) for (j in 1:length(hillterm)) for (k in 1:length(regterm)) {
#     n <- n+1
#     modterms[n,] <- c('elevation',radterm[i],hillterm[j],regterm[k],slterm)
#     modnums[n,] <- c(1,i,j,k,1)
# }
# 
# ## running all possible models on daily basis for four dependent parameters: Tmin, Tmax, RHsat.hrs, VPmax
# RSQelev <- list()
# MSEelev <- list()
# RSQm5x <- list()
# MSEm5x <- list()
# BFm5x <- list()
# RSQm5 <- list()
# MSEm5 <- list()
# 
# vars <- c('Tmin','Tmax','RHsat.hrs','VPmax')
# for (v in 1:length(vars)) {
#     RSQm5[[v]] <- matrix(NA,366,ncol=nModels)
#     MSEm5[[v]] <- matrix(NA,366,ncol=nModels)
#     RSQelev[[v]] <- rep(NA,366)
#     MSEelev[[v]] <- rep(NA,366)
#     RSQm5x[[v]] <- rep(NA,366)
#     MSEm5x[[v]] <- rep(NA,366)
#     BFm5x[[v]] <- rep(NA,366)
# }
# 
# v <- 1
# for (v in 1:length(vars)) {
#     d <- 1
#     for (d in 1:366) {
#         print(c(v,d))
#         dd <- subset(dw,dw$doy==d)
#         dd <- dd[match(dd$siteID,meta$siteID),]
#         dd$yvar <- dd[,vars[v]]
#         dd$yvar[meta$use4Analyses==0] <- NA
#         topo10$dsol <- dlySol[,paste('rad_tot_',days[d],sep='')]
#         
#         #  Base model with elevation only
#         fit1 <- lm(dd$yvar~elevation,data=topo10)
#         RSQelev[[v]][d] <- summary(fit1)$r.sq
#         MSEelev[[v]][d] <- sd(fit1$residuals,na.rm=T)
#         
#         # run all 5 parameter models
#         n <- 0
#         i=1;j=1;k=1
#         for (i in 1:length(radterm)) for (j in 1:length(hillterm)) for (k in 1:length(regterm)) {
#             n <- n+1
#             #print(c(d,i,j,k,n))
#             modterms[n,] <- c('elevation',radterm[i],hillterm[j],regterm[k],slterm)
#             modnums[n,] <- c(1,i,j,k,1)
#             
#             fit2 <- lm(dd$yvar~topo10[,'elevation']+topo10[,radterm[i]]+topo10[,hillterm[j]]+topo10[,regterm[k]]+topo10[,slterm])
#             RSQm5[[v]][d,n] <- summary(fit2)$r.sq
#             MSEm5[[v]][d,n] <- sd(fit2$residuals,na.rm=T)
#         }
#     }
#     
#     ## now extract rsq and mse for the best model, and record id # of best model
#     BFm5x[[v]] <- apply(RSQm5[[v]],1,which.max)
#     for (d in 1:366) {
#         RSQm5x[[v]][d] <- RSQm5[[v]][d,BFm5x[[v]][d]]
#         MSEm5x[[v]][d] <- MSEm5[[v]][d,BFm5x[[v]][d]]
#     }
# }
# 
# pairs(cbind(RSQelev[[1]],RSQelev[[2]],RSQelev[[3]],RSQelev[[4]]))
# pairs(cbind(RSQm5x[[1]],RSQm5x[[2]],RSQm5x[[3]],RSQm5x[[4]]))
# 
# table(BFm5x[[1]],BFm5x[[2]])
# modterms[46,]
# modnums[46,]
# plot(modnums[BFm5x[[2]],3])
# 
# plot(RSQelev[[1]],RSQm5x[[1]])
# plot(RSQelev[[2]],RSQm5x[[2]])
# plot(RSQelev[[3]],RSQm5x[[3]])
# plot(RSQelev[[4]],RSQm5x[[4]])
# 
# plot(MSEelev[[1]],MSEm5x[[1]]);abline(0,1)
# plot(MSEelev[[2]],MSEm5x[[2]]);abline(0,1)
# plot(MSEelev[[3]],MSEm5x[[3]]);abline(0,1)
# plot(MSEelev[[4]],MSEm5x[[4]]);abline(0,1)
# 
# head(dlySummary)
# boxplot(MSEm5x[[2]]~dlySummary$sommax)
# boxplot(MSEm5x[[1]]~dlySummary$sommin)
# 
# plot(RSQm5x[[1]]~dlySummary$Kps.mean_hPa)
# plot(RSQm5x[[2]]~dlySummary$Kps.mean_hPa)
# plot(RSQm5x[[3]]~dlySummary$Kps.mean_hPa)
# plot(RSQm5x[[4]]~dlySummary$Kps.mean_hPa)
# 
# plot(RSQm5x[[1]]~dlySummary$Kwspd.mean_m.s)
# plot(RSQm5x[[2]]~dlySummary$Kwspd.mean_m.s)
# plot(RSQm5x[[3]]~dlySummary$Kwspd.mean_m.s)
# plot(RSQm5x[[4]]~dlySummary$Kwspd.mean_m.s)
# 
# plot(RSQm5x[[1]]~dlySummary$Tmin)
# plot(RSQm5x[[2]]~dlySummary$Tmax)
# plot(RSQm5x[[3]]~dlySummary$RHsat.hrs)
# plot(RSQm5x[[4]]~dlySummary$VPmax)
# 
# plot(RSQm5x[[1]]~I(dlySummary$Tmax - dlySummary$Tmin))
# plot(RSQm5x[[2]]~I(dlySummary$Tmax - dlySummary$Tmin))
# plot(RSQm5x[[3]]~I(dlySummary$Tmax - dlySummary$Tmin))
# plot(RSQm5x[[4]]~I(dlySummary$Tmax - dlySummary$Tmin))
# 
# plot(I(dlySummary$Tmax - dlySummary$Tmin)~dlySummary$Tmin)
# plot(I(dlySummary$Tmax - dlySummary$Tmin)~dlySummary$Tmax)
# plot(I(dlySummary$Tmax - dlySummary$Tmin)~dlySummary$RHsat.hrs)
# plot(I(dlySummary$Tmax - dlySummary$Tmin)~dlySummary$VPmax)

```

