---
title: "TableMt_2012"
author: "D. Ackerly, J. Slingsby"
date: "February 14, 2016"
output: html_document
---

In 2012, we placed 96 dataloggers recording temperature and humidity across Table Mt. and Silvermine. The loggers were stratified across gradients of elevation, slope, exposure and hillslope position (hilltop, valleybottom), and distance from both False Bay and the Atlantic. In previous analyses, the data has been analyzed to extract diurnal Tmin, Tmax, VPmax (maximum daily vapor pressure), and RHsat.hrs (hours per day with relative humidity above 95%).

In preliminary analyses, we found that during periods of rapidly changing weather, the Tmin for one day might occur at midnight (if the previous night had been warmer) and then another Tmin would be recorded in the early morning of the next day. Similar patterns were observed for Tmax. In order to minimize this problem, we calculated Tmin values for each day in the 24 hours between 1500h the prior day and 1500h on a given day. Similarly, we calculated Tmax in the 24 hours between 0600h on the day in question and 0600h on the following day. The time at which Tmin and Tmax were observed is also recorded, with negative values for Tmin indicating the prior day, and values >24 for Tmax indicating the next day.

Tmin and Tmax may not be observed at the same time at different locations, which may reduce the apparent environmental lapse rates. For a more direct analysis of lapse rates, we also examine temperature at 0600h and 1400h, which are approximately when Tmin and Tmax, respectively, are most often observed. See 'sandbox/extract-6-14h-temps.R' for extraction script for those values.

All of these analyses and the summary data are available at: 
https://github.com/dackerly/table-mt-microclimate-2012.git

General summary statistics about the sites and daily weather values can be found in 'manuscript-summary-stats.r'. 

The goal of these analyses is to examine spatial variation in daily weather values, running models for each day through 2012. The initial question is whether the important spatial variables vary with season, but as shown below a key result is enormous day to day variation in the ability to explain spatial variation in weather variables. I'm still searching for explanations in regional weather or other factors to understand why the models differ so much from day to day.

```{r}
#----------------#
# LOAD LIBRARIES #
#----------------#
rm(list=ls())
library(raster)
library(RColorBrewer)

```
'meta' contains metadata on the 96 sites. 'topo10' and 'topo30' have a series of topographic features sampled from 10m and 30m dem rasters, respectively, based on lat/long positions of each site. SiteID is the unique name for each site. 

Some of the variables in 'meta':
* elevation, slope, and aspect: measured in the field - these are generally not used, and instead we use a consistent set of values taken from the 10m dem.
* use4Analyses, use4ClimateSummaries: sites excluded due to incomplete data (see ms for criteria)

Variables in topo10 and topo30:
* d2at, d2fb, d2cs: distance to atlantic, distance to false bay, distance to coast = min(d2at,d2fb)
* rad080, rad172, rad355: solar radiation at summer and winter solstice, and equinox. Not generally used as we have diurnal radiation at each site
* thl0, thl315, thl337: Topographic heat load as defined by McCune and Keon 2002 JVS, with the axis for calculation of the 'folded aspect' (and thus the maximum heat load) at 0, 337 or 315 degrees. The logic is that in the S Hem, NW facing slopes would be hotter than N facing, due to higher afternoon temperatures.
* plow050, 125, 250, 500: Percent lower pixels - the percent of pixels surrounding a focal pixel with a lower elevation, in radii of 50, 125, 250 or 500 m. A measure of potential for cold air to flow away from a spot. Higher values indicate hilltops
* tpi050, 125, 250, 500: Elevation of a focal pixel - mean elevation in the specified radius. Another measure of hillslope position, with positive values for hilltops, negative for valley bottoms.

```{r}
#----------------#
# LOAD DATA      #
#----------------#
meta <- read.csv('data/csv_masters/location_meta.csv',as.is=T)
dim(meta)
head(meta)

topo10 <- read.csv('data/csv_masters/topo10.csv',as.is=T)
dim(topo10)
names(topo10)
 
topo30 <- read.csv('data/csv_masters/topo30.csv',as.is=T)
dim(topo30)
names(topo30)
```

'dw' has all daily weather values for all sites and days. The dlySummary contains summaries of weather data on each day - average weather across sites. dlySol has diurnal radiation at each site for each day of the year, from Adam Wilson's analyses on a 30m dem.
```{r}
dw <- read.csv('data/csv_masters/2012daily.csv',as.is=T)
head(dw)

dlySummary <- read.csv('data/csv_outfiles/dlySummary.csv',as.is=T)
head(dlySummary)

dlySol <- readRDS('data/Rdata/dlySol.Rdata')
dlySol[1:6,1:6]
```

Before fitting any spatial models of weather variation, we want to check for correlations and spatial autocorrelation of predictors. Based on the topo10 variables, here are all variables with pairwise correlations R2 > 0.5. Commented code can be run to see more detail. As expected, radiation load variables are mostly correlated, and hillslope position variables are correlated. Below, we run alternative models swapping these in and out, but not using them in the same model.

```{r}

# cor(topo10[,-c(1:2)],use='pair')
# pairs(topo10[,-c(1:2)])
# names(topo10)
# 
## find correlations > 0.5
i <- 3; j <- 4;
for (i in 3:(ncol(topo10)-1)) for (j in (i+1):ncol(topo10)) {
    rr <- cor(topo10[,i],topo10[,j],use='pair')
    if (abs(rr)>sqrt(0.5)) print(c(names(topo10)[c(i,j)],round(rr,3)))
}
# 
# # distance from Atlantic and False Bay, borderline
# # Ran some models swapping these two variables in and out, and found false bay on average increased R2 more, and overall model R2 are significantly correlated so when one works, either works overall. Based on this, all models use Distance to False Bay.
# cor(topo10[,c('d2at','d2fb')],use='pair') # R = -0.65
# 
# # equinox and winter solstice radiation
# cor(topo10[,c('rad080','rad172')],use='pair') # R = 0.95
# 
# # all tpi variables with each other across scales, and plow with each other across scales; elevation weakly correlated with tpi and plow
# cor(topo10[,c('elevation','tpi050','tpi125','tpi250','tpi500','plow050','plow125','plow250','plow500')],use='pair') # R >= 0.89
# cor(topo30[,c('elevation','tpi050','tpi125','tpi250','tpi500','plow050','plow125','plow250','plow500')],use='pair') # R >= 0.89
# 
```

How does the Topographic Heat Load index, calculated with different aspect orientations, compare to daily radiation load from a solar model? Maximum values are near the equinoxes, with a minimum at summer solstice and a dip at winter solstice. thl315 has lower correlation as it is furthest from north-facing, so may work for temperature due to hotter afternoon air temp, but is not as strongly correlated with solar radiation per se.
```{r}
# # How does daily solar radiation compare to topographic heat load
# # correlation max at 0.7 on day 48 (Feb 17) and around day 295 (Oct 21)
# # minimum mid-summer solstice
# head(dlySol)
# head(topo10)
solVthl <- c()
i=1
for (i in 1:366) solVthl[i] <- cor(topo10$thl0,dlySol[,i],use='pair')
plot(1:366,solVthl,type='l')   

for (i in 1:366) solVthl[i] <- cor(topo10$thl315,dlySol[,i],use='pair')
lines(1:366,solVthl,col='red')

for (i in 1:366) solVthl[i] <- cor(topo10$thl337,dlySol[,i],use='pair')
lines(1:366,solVthl,col='blue')   

```

The following shows scatterplots of pairwise differences matrices for spatial distance vs. each environmental factor, and the accompanying Mantel test. The plots that are commented out are not significant for spatial correlation, and plots that are significant are illustrated. The random string in the box is an identifier to link this code to the results in the manuscript. Site 32 (Kirstenbosch Race Track) is removed from analysis.

As expected, distance from ocean variables are strongly spatially correlated. This raises questions about their interpretation in the models, but they are generally their for the purpose of capturing spatial gradients, so that could be okay. Elevation and slope are also spatially structured, though the scatterplots suggests less concern. Exposure and hillslope positions are not correlated, indicating we did a good job stratifying across the region.

```{r}
# 
# #--------------------------------------------#
# # SPATIAL VARIATION IN TOPOGRAPHIC VARIABLES #
# # a6STycTJVjqM8vRd8V8h                       #
# #--------------------------------------------#
library(ade4)
# check matrices are aligned
all(meta$siteID==topo10$siteID)

#head(meta)
sdist <- dist(meta[-32,c('UTM.east','UTM.north')])
hist(sdist,main='Histogram of distances between sites (m)')

selev <- dist(meta$elevation[-32])
plot(sdist,selev)
mantel.rtest(sdist,selev,nrepet=999)

sd2at <- dist(topo10$d2at[-32])
plot(sdist,sd2at)
mantel.rtest(sdist,sd2at,nrepet=999)

sd2fb <- dist(topo10$d2fb[-32])
plot(sdist,sd2fb)
mantel.rtest(sdist,sd2fb,nrepet=999)

plot(selev,sd2fb)
mantel.rtest(selev,sd2fb,nrepet=999)

sd2cs <- dist(topo10$d2cs[-32])
plot(sdist,sd2cs)
mantel.rtest(sdist,sd2cs,nrepet=999)

dslope <- dist(topo10$slope[-32])
plot(sdist,dslope)
mantel.rtest(sdist,dslope,nrepet=999)

dtpi50 <- dist(topo10$tpi050[-32])
#plot(sdist,dtpi50)
mantel.rtest(sdist,dtpi50,nrepet=999)

drad080 <- dist(topo10$rad080[-32])
#plot(sdist,drad080)
mantel.rtest(sdist,drad080,nrepet=999)

dthl <- dist(topo10$thl0[-32])
#plot(sdist,dthl)
mantel.rtest(sdist,dthl,nrepet=999)
```

```{r echo=F}
# DEPRECATED CODE
# #-----------------------#
# # LOAD DLY WEATHER DATA #
# # Leaving commented out for now, as this has been replaced by 'dw' data.frame #
# #-----------------------#
# 
# cbday <- readRDS('data/Rdata/cbday.Rdata')
# length(cbday)
# head(cbday[[1]])
# dim(cbday[[1]])
# 
```

```{r}
# #-------------------------------------------------#
# # MAKE 3 CHAR DAYS VAR, for dlySol variable names #
# #-------------------------------------------------#
days <- as.character(1:366)
days[nchar(days)==1] <- paste('00',days[nchar(days)==1],sep='')
days[nchar(days)==2] <- paste('0',days[nchar(days)==2],sep='')
```

Okay, the modeling strategy is a bit brute force now! First set up four variables that have alternative terms for equivalent topographic variables: radterm = solar radiation; hillterm = topographic hillslope position; regterm = regional position relative to ocean; slterm = slope (only one variable). Plus there's elevation in every model. Then set up modterms and modnums which contain the names and numbers of each item for all 120 combinations (5 radterm * 8 hillterm * 3 regterm). And finally set up a bunch of lists which will have four items each, holding results for Tmin, Tmax, RHsat.hrs, and VPmax; each item in the list will be a 366 row matrix with model results for each day

```{R}
#-------------------------------#
# DAILY WEATHER MODELS          #
# OHkNt68RXvbTk4jHfczb          #
#-------------------------------#

# Run through daily models testing all combinations of elevation, radiation, hillslope, and regional position terms
radterm <- c('dsol','rad080','rad172','rad355','thl315','thl337','thl0')
hillterm <- c('tpi050','tpi125','tpi250','tpi500','plow050','plow125','plow250','plow500')
regterm <- c('d2fb','d2at','d2cs')
slterm <- 'slope'
nModels <- length(radterm)*length(hillterm)*length(regterm)

modterms <- matrix(NA,nModels,5)
modnums <- matrix(NA,nModels,5)
n <- 0
for (i in 1:length(radterm)) for (j in 1:length(hillterm)) for (k in 1:length(regterm)) {
    n <- n+1
    modterms[n,] <- c('elevation',radterm[i],hillterm[j],regterm[k],slterm)
    modnums[n,] <- c(1,i,j,k,1)
}

## running all possible models on daily basis for six dependent parameters: Tmin, Tmax, T06, T14, RHsat.hrs, VPmax
RSQelev <- list() # r-squared, elevation only model
MSEelev <- list() # mean square error, elevation only model
RSQm5x <- list() # highest r-squared, 5 term model
MSEm5x <- list() # mean square error for max r-squared 5 term model
BFm5x <- list() # best-fit model: number from 1:120 indicating which had highest r-squared
RSQm5 <- list() # matrix of 120 r-squared for all 5 term models
MSEm5 <- list() # matrix of 120 mean square errors for all 5 term models
dAICx <- list() # delta AIC from drop1 function for bf model
rSlpx <- list() # regression slopes from bg model

vars <- c('Tmin','Tmax','T06','T14','RHsat.hrs','VPmax')
for (v in 1:length(vars)) {
    RSQm5[[v]] <- matrix(NA,366,ncol=nModels)
    MSEm5[[v]] <- matrix(NA,366,ncol=nModels)
    RSQelev[[v]] <- rep(NA,366)
    MSEelev[[v]] <- rep(NA,366)
    RSQm5x[[v]] <- rep(NA,366)
    MSEm5x[[v]] <- rep(NA,366)
    BFm5x[[v]] <- rep(NA,366)
    dAICx[[v]] <- matrix(NA,366,5)
    rSlpx[[v]] <- matrix(NA,366,5)
}
```

Now, run through the four weather variables (v loop) and the 366 days (d loop). Each time through, subset the weather data, and insert the appropriate daily solar radiation from the dlySol data.frame into topo10$dsol. Fit one model with elevation only, and record results. Then fit all 120 possible 5 terms models (i,j,k, loop), saving r-squared and MSE. After running all of them, extract best fit and corresponding r-sq and MSE.

```{r}
v <- 1
for (v in 1:length(vars)) {
    d <- 1
    for (d in 1:366) {
        #print(c(v,d))
        dd <- subset(dw,dw$doy==d)
        dd <- dd[match(dd$siteID,meta$siteID),]
        dd$yvar <- dd[,vars[v]]
        dd$yvar[meta$use4Analyses==0] <- NA
        topo10$dsol <- dlySol[,paste('rad_tot_',days[d],sep='')]
        
        #  Base model with elevation only
        fit1 <- lm(dd$yvar~elevation,data=topo10)
        RSQelev[[v]][d] <- summary(fit1)$r.sq
        MSEelev[[v]][d] <- sd(fit1$residuals,na.rm=T)
        
        # run all 5 parameter models
        n <- 0
        i=1;j=1;k=1
        for (i in 1:length(radterm)) for (j in 1:length(hillterm)) for (k in 1:length(regterm)) {
            n <- n+1
            #print(c(d,i,j,k,n))

            fit2 <- lm(dd$yvar~topo10[,'elevation']+topo10[,radterm[i]]+topo10[,hillterm[j]]+topo10[,regterm[k]]+topo10[,slterm])
            RSQm5[[v]][d,n] <- summary(fit2)$r.sq
            MSEm5[[v]][d,n] <- sd(fit2$residuals,na.rm=T)
#             if (n==1) {
#                 mrs <- summary(fit2)$r.sq
#                 d1 <- coefficients(drop1(fit2))
#                 dAICx[[v]][d,] <- drop1(fit2)$AIC[-1]
#             } else {
#                 if (summary(fit2)$r.sq > mrs) {
#                     mrs <- summary(fit2)$r.sq
#                     d1 <- coefficients(drop1(fit2))
#                     dAICx[[v]][d,] <- drop1(fit2)$AIC[-1]
#                 }
#             }
        }
        ## now extract rsq and mse for the best model, and record id # of best model
        BFm5x[[v]][d] <- which.max(RSQm5[[v]][d,])
        RSQm5x[[v]][d] <- RSQm5[[v]][d,BFm5x[[v]][d]]
        MSEm5x[[v]][d] <- MSEm5[[v]][d,BFm5x[[v]][d]]
        
        # rerun best model to extract dAIC values
        i <- modnums[BFm5x[[v]][d],2]
        j <- modnums[BFm5x[[v]][d],3]
        k <- modnums[BFm5x[[v]][d],4]
        fit2 <- lm(dd$yvar~topo10[,'elevation']+topo10[,radterm[i]]+topo10[,hillterm[j]]+topo10[,regterm[k]]+topo10[,slterm])
        d1 <- drop1(fit2)$AIC
        dAICx[[v]][d,] <- d1[-1]-d1[1]
        rSlpx[[v]][d,] <- fit2$coefficients[-1]
    }
}

saveRDS(list(RSQelev,MSEelev,RSQm5,MSEm5,RSQm5x,MSEm5x,BFm5x,dAICx,rSlpx,modterms,modnums),file='data/Rdata/M5out.Rdata')
```
